---
title: "JAGS Code for *Shifts in elevational distributions of montane birds in an arid ecosystem*"
output: html_document
author: Martha W. Zillig 
---

This code includes model formulations for a single species multi-season occupancy model examining elevational shifts. 

First, load the data and required packages:

```{r}
library(tidyverse)
library(abind)
library(jagsUI)
library(boot)

#data for Brewer's Sparrow in the central Great Basin sites:

data <- readRDS("brsp_model_data.RDS")
brsp_matrix <- readRDS("brsp_matrix.RDS")
```

Parameterization for JAGS model 

```{r}
n_obs <- 37 #number of unique observers 

a.vec <- rep(1,n_obs)

data_q1 <- data

C <- brsp_matrix

Zst <- apply(C, 1, max, na.rm =TRUE)

inits <- function()list(Z = Zst, alpha0 = runif(1), beta0 = dnorm(1))

params <- c("betaT", "theta1", "theta2", "theta3", "theta4", "alpha0", "beta0", "alpha1", "alpha2", "bp.value", "bp.value.n", "ind")
```

JAGS model for *full elevational gradient* in the Central Great Basin sites: 

```{r}

cat("

model { 

  for(i in 1:n_point){
    alpha1[i] ~ dnorm(0, sigma.y)
  }
  
  for(i in 1:n_canyon){
    alpha2[i] ~ dnorm(0, sigma.c)
  }
    for(i in 1:n_obs){
    theta3[i] ~ dnorm(0, tau.o)
  }
  
  for(i in 1:M){   #missing yday values
    for(j in 1:J){
    yday[i,j] ~ dnorm(mu.yday, tau.yday)
    }
  }
  
  for(i in 1:M){   #missing time values
    for(j in 1:J){
    time[i,j] ~ dnorm(mu.time, tau.time)
    }
  }
    for(i in 1:M){ #missing observer values 
    for (a in 1:3){
      obs_num[i,a] ~ dcat(b[1:n_obs])  
    }
  }
  
  #Indicator variable priors
  
  for (j in 1:Q) {
    ind[j]~dbern(0.5)
    betaT[j]~dnorm(0,tauT)
    beta[j]<-ind[j]*betaT[j]
  }
  
  tauT ~ dgamma(0.1,0.1)
  
  for(i in 1:M) 
  {
    Z[i] ~ dbern(lambda[i])
    
    logit(lambda[i]) <- alpha0 + inprod(X[i,],beta) + alpha1[point[i]] + alpha2[canyon[i]]
    
    n.new[i]~ dbern(lambda[i]) #calculation of bayes p value
    res.n[i] <- (sqrt(Z[i]) - sqrt(lambda[i]))^2
    res.n.new[i] <- (sqrt(n.new[i]) - sqrt(lambda[i]))^2
    
    
 for (j in 1:J) {
 
      C[i,j] ~ dbin(p[i,j], Z[i])
      logit(p[i,j]) <- beta0  + theta1*yday[i,j] + theta2*time[i,j] +theta4*pow((time[i,j]),2) + theta3[obs_num[i,j]]
      
      y.new[i,j]~ dbin(p[i,j], Z[i]) #calculation of bayes p values 
      res[i,j] <- (sqrt(C[i,j])-sqrt(p[i,j]*Z[i]))^2
      res.new[i,j] <- (sqrt(y.new[i,j])-sqrt(p[i,j]*Z[i]))^2
      
    }
  }
  
  #Bayesian p-value calculations 
  
  total.res.n <- sum(res.n[])
  total.resnew.n <- sum(res.n.new[])
  bp.value.n <- total.res.n>total.resnew.n
  
  total.res <- sum(res[,])
  total.resnew <- sum(res.new[,])
  bp.value <- total.res>total.resnew
  
  beta0 ~ dlogis(0,1)
  alpha0 ~ dnorm(0,1)
  mu.yday ~ dnorm(0,.001)
  tau.yday <- pow(sd.yday, -2)
  sd.yday ~ dunif(0,100)
  mu.time ~ dnorm(0,.001)
  tau.time <- pow(sd.time, -2)
  sd.time ~ dunif(0,100)
  sigma.y ~ dunif(0,1)
  sigma.c ~ dunif(0,1)
  theta1 ~ dnorm(0,1)
  theta2 ~ dnorm(0,1)
  theta4 ~ dnorm(0,1)
  tau.o ~ dunif(0,1)
  b[1:n_obs] ~ ddirch(a.vec)
  

}
", file = "occ_model_ch2.txt")

#Model run time is quite long (~9 hours), we recommend reducing number of iterations for exploration. 

#out_brsp <- jags(data = data, model.file = "occ_model_ch2.txt", inits = inits, parameters.to.save = params, n.chains = 3, n.iter = 50000, n.burnin = 10000, n.adapt = 10000, n.thin = 20)


```

To examine potential elevational shifts at range edges, we used the same model, but used WAIC instead of indicator variable selection. We compared four different models using WAIC: an intercept only model, a model with all covaraites included, a model with year, elevation, and the interaction of year and elevation included, and a model with only biotic variables included (spring precip., spring temperature, winter precip.,winter temperature, NDVI). 

The model with all covariates included for the upper 25% of the elevational gradient is shown below.

First, load the data for American Robin in the Central Great Basin sites. This is data for the top 25% of the elevational gradient:

```{r}
amro_q1_matrix <-readRDS("amro_matrix_q1.RDS")
amro_data <- readRDS("amro_dataQ1.RDS")

#JAGS model parameterization

n_obs <- 37 #number of unique observers 

a.vec <- rep(1,n_obs)

C <- amro_q1_matrix

data1 <- amro_data

Zst <- apply(C, 1, max, na.rm =TRUE)

inits <- function()list(Z = Zst, alpha0 = runif(1), beta0 = dnorm(1))

params <- c("beta1", "beta2", "beta3", "beta4", "beta5", "beta6", "beta7", "theta1", "theta2", "theta3", "theta4", "alpha0", "beta0", "alpha1", "alpha2", "bp.value", "bp.value.n", "ld")

#JAGS Model for upper and lower 25% of elevational gradient: 
#note only change is no indicator variables and calculation of WAIC

cat("

model { 

  for(i in 1:n_point){
    alpha1[i] ~ dnorm(0, sigma.y)
  }
  
  for(i in 1:n_canyon){
    alpha2[i] ~ dnorm(0, sigma.c)
  }
    for(i in 1:n_obs){
    theta3[i] ~ dnorm(0, tau.o)
  }
  
  for(i in 1:M){   #missing yday values
    for(j in 1:J){
    yday[i,j] ~ dnorm(mu.yday, tau.yday)
    }
  }
  
  for(i in 1:M){   #missing time values
    for(j in 1:J){
    time[i,j] ~ dnorm(mu.time, tau.time)
    }
  }
    for(i in 1:M){ #missing observer values 
    for (a in 1:3){
      obs_num[i,a] ~ dcat(b[1:n_obs])  
    }
  }
  
  
  for(i in 1:M) 
  {
    Z[i] ~ dbern(lambda[i])
    
    logit(lambda[i]) <- alpha0 + beta1*year[i] + beta2*elev[i] + beta3*interact[i] + beta4*spring_ppt[i] + beta5*spring_temp[i] + beta6*winter_ppt[i] + beta7*max_ndvi[i] + alpha1[point[i]] + alpha2[canyon[i]]
    
    n.new[i]~ dbern(lambda[i]) #calculation of bayes p value
    res.n[i] <- (sqrt(Z[i]) - sqrt(lambda[i]))^2
    res.n.new[i] <- (sqrt(n.new[i]) - sqrt(lambda[i]))^2
    
    ld[i] <- logdensity.bern(Z[i], lambda[i]) #calculation of log density to use in WAIC calculation 
    
 for (j in 1:J) {
 
      C[i,j] ~ dbin(p[i,j], Z[i])
      logit(p[i,j]) <- beta0  + theta1*yday[i,j] + theta2*time[i,j] +theta4*pow((time[i,j]),2) + theta3[obs_num[i,j]]
      
      y.new[i,j]~ dbin(p[i,j], Z[i]) #calculation of bayes p values 
      res[i,j] <- (sqrt(C[i,j])-sqrt(p[i,j]*Z[i]))^2
      res.new[i,j] <- (sqrt(y.new[i,j])-sqrt(p[i,j]*Z[i]))^2
      

    }
  }
  
  
  total.res.n <- sum(res.n[])
  total.resnew.n <- sum(res.n.new[])
  bp.value.n <- total.res.n>total.resnew.n
  
  total.res <- sum(res[,])
  total.resnew <- sum(res.new[,])
  bp.value <- total.res>total.resnew
  
  beta0 ~ dlogis(0,1)
  alpha0 ~ dnorm(0,1)
  mu.yday ~ dnorm(0,.001)
  tau.yday <- pow(sd.yday, -2)
  sd.yday ~ dunif(0,100)
  mu.time ~ dnorm(0,.001)
  tau.time <- pow(sd.time, -2)
  sd.time ~ dunif(0,100)
  sigma.y ~ dunif(0,1)
  sigma.c ~ dunif(0,1)
  theta1 ~ dnorm(0,1)
  theta2 ~ dnorm(0,1)
  theta4 ~ dnorm(0,1)
  tau.o ~ dunif(0,1)
  beta1 ~ dnorm(0,1)
  beta2 ~ dnorm(0,1)
  beta3 ~ dnorm(0,1)
  beta4 ~ dnorm(0,1)
  beta5 ~ dnorm(0,1)
  beta6 ~ dnorm(0,1)
  beta7 ~ dnorm(0,1)
  b[1:n_obs] ~ ddirch(a.vec)
  

}
", file = "q_model_full.txt")

#Run model: 

#out_amro_full <- jags(data = data1, model.file = "q_model_full.txt", inits = inits, parameters.to.save = params, n.chains = 3, n.iter = 50000, n.burnin = 10000, n.adapt = 10000, n.thin = 20)

```




