---
title: "JAGS Code for *Shifts in elevational distributions of montane birds in an arid ecosystem*"
output: html_document
author: Martha W. Zillig 
---

This code includes model formulations for a single species multi-season occupancy model examining elevational shifts. 

First, load the data and required packages:

```{r}
library(tidyverse)
library(abind)
library(jagsUI)
library(boot)

#data for Brewer's Sparrow in the central Great Basin sites:

data_brsp <- readRDS("brsp_model_data.RDS")
brsp_matrix <- readRDS("brsp_matrix.RDS")
```

Parameterization for JAGS model 

```{r}
n_obs <- 37 #number of unique observers 

a.vec <- rep(1,n_obs)

C <- brsp_matrix

data <- data_brsp

Zst <- apply(C, 1, max, na.rm =TRUE)

inits <- function()list(Z = Zst, alpha0 = runif(1), beta0 = dnorm(1))

params <- c("beta1", "beta2", "beta3", "beta4", "beta5", "beta6", "beta7", "theta1", "theta2", "theta3", "theta4", "alpha0", "beta0", "alpha1", "alpha2", "bp.value", "bp.value.n", "ld")

```
JAGS MODEL

Below we include code for one of the four models examined with WAIC. This is the "full" model with all covariates included. The other three models tested are "intercept only" (no fixed effects on the abundance or detection sub-models),  the "abiotic" model (fixed effects of year, elevation, and the interaction of year and elevation included on the abundance sub-model), and the "biotic" model (fixed effects of spring precipitation, spring temperature, winter precipitation, winter temperature and NDVI included on the abundance sub-model). 

cat("

model { 

  for(i in 1:n_point){ #random effect of survey point 
    alpha1[i] ~ dnorm(0, tau.y)
  }
  
  for(i in 1:n_canyon){ #random effect of canyon 
    alpha2[i] ~ dnorm(0, tau.c)
  }
    for(i in 1:n_obs){ #random effect of observer 
    theta3[i] ~ dnorm(0, tau.o)
  }
  
  for(i in 1:M){   #missing yday value
    for(j in 1:J){
    yday[i,j] ~ dnorm(mu.yday, tau.yday)
    }
  }
  
  for(i in 1:M){   #missing time value
    for(j in 1:J){
    time[i,j] ~ dnorm(mu.time, tau.time)
    }
  }
    for(i in 1:M){ #missing observer values 
    for (a in 1:3){
      obs_num[i,a] ~ dcat(b[1:n_obs])  
    }
  }
  
  
  for(i in 1:M) 
  {
    Z[i] ~ dbern(lambda[i])
    
    logit(lambda[i]) <-  alpha0 + beta1*year[i] + beta2*elev[i] + beta3*interact[i] + beta4*spring_ppt[i] + beta5*spring_temp[i] + beta6*winter_ppt[i] + beta7*max_ndvi[i] + alpha1[point[i]] + alpha2[canyon[i]]
    
    n.new[i]~ dbern(lambda[i]) #calculation of bayes p value
    res.n[i] <- (sqrt(Z[i]) - sqrt(lambda[i]))^2
    res.n.new[i] <- (sqrt(n.new[i]) - sqrt(lambda[i]))^2
    
    ld[i] <- logdensity.bern(Z[i], lambda[i]) #to calculate WAIC score 
    
    
 for (j in 1:J) {
 
      C[i,j] ~ dbin(p[i,j], Z[i])
      logit(p[i,j]) <- beta0  + theta1*yday[i,j] + theta2*time[i,j] +theta4*pow((time[i,j]),2) + theta3[obs_num[i,j]]
      
      y.new[i,j]~ dbin(p[i,j], Z[i]) #calculation of bayes p values 
      res[i,j] <- (sqrt(C[i,j])-sqrt(p[i,j]*Z[i]))^2
      res.new[i,j] <- (sqrt(y.new[i,j])-sqrt(p[i,j]*Z[i]))^2
      
    }
  }
  
  
  total.res.n <- sum(res.n[])
  total.resnew.n <- sum(res.n.new[])
  bp.value.n <- total.res.n>total.resnew.n
  
  total.res <- sum(res[,])
  total.resnew <- sum(res.new[,])
  bp.value <- total.res>total.resnew
  
  beta0 ~ dt(0, 1, 6) 
  alpha0 ~ dt(0, 1, 6)
  mu.yday ~ dnorm(0,.001)
  tau.yday <- pow(sd.yday, -2)
  sd.yday ~ dnorm(0,1) T(0,)
  mu.time ~ dnorm(0,.001)
  tau.time <- pow(sd.time, -2)
  sd.time ~ dt(0, 1, 6) T(0,)
  tau.c <- pow(sigma.c, -2)
  tau.y <- pow(sigma.y, -2)
  tau.o <- pow(sigma.o, -2)
  sigma.o ~ dt(0, 1, 6) T(0,)
  sigma.y ~ dt(0, 1, 6) T(0,)
  sigma.c ~ dt(0, 1, 6) T(0,)
  theta1 ~ dt(0, 1, 6)
  theta2 ~ dt(0, 1, 6)
  theta4 ~ dt(0, 1, 6)
  beta1 ~ dt(0, 1, 6)
  beta2 ~ dt(0, 1, 6)
  beta3 ~ dt(0, 1, 6)
  beta4 ~ dt(0, 1, 6)
  beta5 ~ dt(0, 1, 6)
  beta6 ~ dt(0, 1, 6)
  beta7 ~ dt(0, 1, 6)
  b[1:n_obs] ~ ddirch(a.vec)
  

}
", file = "occ_model_ch2.txt")

#Run model with below code. Run time is approx 2 hours. 

out_brsp <- jags(data = data, model.file = "occ_model_ch2.txt", inits = inits, parameters.to.save = params, n.chains = 3, n.iter = 60000, n.burnin = 10000, n.adapt = 10000, n.thin = 10)

```

